---
title: "Bioinformatics_Plots"
author: "Jaret Arnold"
date: "2025-04-20"
output: html_document
---

```{r}
#imports and file load
library(tidyverse)
library(ggplot2)

hybrids <- read.csv("../data/orchidroots_cattleya_hybrids.csv")
species <- read.csv("../data/orchidroots_cattleya_species.csv")

summary(hybrids)
summary(species)

```


```{r}
#Data Cleaning

# Remove subsp. and var. from Species entries 
species_filtered <- species[grepl("subsp\\. |\\svar\\.", species$Binomial) == FALSE, ]


# Remove any hybrids which are just synonyms 
hybrids_filtered <- hybrids[!grepl("Synonym of", hybrids$Parentage), ]

# Filter out the year 2025
hybrids_filtered <- hybrids_filtered[hybrids_filtered$Year != 2025, ]

```


```{r}
################################################################################
####################################hybrid complexity vs time w/ annotation#####
# Define Corsage Era
corsage_era <- data.frame(
  xmin = 1920,
  xmax = 1950,
  ymin = -Inf,
  ymax = Inf,
  era = "Cattleya Corsage Era"
)

# Map labels safely and directly in the plotting dataframe
df_hybrid_trend <- df_hybrid_trend %>%
  mutate(HybridTypeLabel = recode(HybridType,
    "Primary Hybrid" = "Primary (species × species)",
    "Complex Hybrid" = "Complex (hybrid × species)",
    "Highly Complex Hybrid" = "Highly Complex (hybrid × hybrid)"
  ))

# Combined Plot
ggplot() +
  # Shaded corsage era
  geom_rect(data = corsage_era, 
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = era),
            alpha = 0.2, inherit.aes = FALSE) +
  
  # Hybrid trend lines and points
  geom_line(data = df_hybrid_trend, 
            aes(x = Year, y = Count, color = HybridTypeLabel), size = 1) +
  geom_point(data = df_hybrid_trend, 
             aes(x = Year, y = Count, color = HybridTypeLabel), size = 1.5) +

  # Custom fill and color scales
  scale_fill_manual(name = NULL, values = c("Cattleya Corsage Era" = "purple")) +

  labs(
    title = "Hybridization Trends Over Time",
    x = "Year",
    y = "Count",
    color = "Hybrid Type"
  ) +

  theme_minimal() +

  # Ensure both legends appear clearly
  guides(
    fill = guide_legend(override.aes = list(alpha = 0.2)),
    color = guide_legend(override.aes = list(size = 2))
  )

#####################################################Year vs Hybrid graph######

year_counts <- table(hybrids_filtered$Year)
df_years <- as.data.frame(year_counts)
colnames(df_years) <- c("Year", "Count")
df_years$Year <- as.numeric(as.character(df_years$Year))

ggplot(df_years, aes(x = as.numeric(Year), y = Count, group = 1)) +
  geom_line(color = "blue", size = 1) +   
  geom_point(color = "red", size = 2) + 
  #geom_smooth(method = "loess", se = FALSE,
  #            color = "black", linetype = "dashed") +
  labs(title = "Number of Cattleya Hybrids per Year", x = "Year", y = "Number of Hybrids") +
  theme_minimal() +
  # Show only every ten years (starting at 1850) for readability 
  scale_x_continuous(breaks = seq(1850, max(df_years$Year, na.rm = TRUE), by = 10))
 #trendline included above for visualization

#########################################################
#################Ancestor vs Year distrib#################


df_clean <- hybrids_filtered %>%
  filter(!is.na(Year), !is.na(Ancestors),
         is.finite(Year), is.finite(Ancestors))


ggplot(df_clean, aes(x = Year, y = Ancestors)) +
  geom_boxplot(aes(group = Year), fill = "lightblue") +
  geom_smooth(method = "loess", se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Number of Ancestors by Year",
       x = "Year", y = "Number of Ancestors") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1850, max(hybrids_filtered$Year, na.rm = TRUE), by = 10))

############################hybrids cumulative vs year##########################

hybrids_filtered %>%
  filter(!is.na(Year)) %>%
  count(Year) %>%
  arrange(Year) %>%
  mutate(cumulative_hybrids = cumsum(n)) %>%
  ggplot(aes(x = Year, y = cumulative_hybrids)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 1.5, color = "steelblue") +
  labs(title = "Cumulative Number of Registered Hybrids Over Time",
       x = "Year", y = "Cumulative Hybrid Count") +
  theme_minimal()



```


##############################################################################
###########################Test for phylogenetic Signal######################
##############################################################################

library(ape)

# Open a new plotting window (Windows only)
windows(width = 11, height = 8.5)

# Set margins
par(mar = c(5, 4, 4, 12))

# Plot the subgeneric tree
plot(tree_pruned, cex = 0.8, font = 3, no.margin = TRUE)

#############################################################
#############################################################

#Cleaning:
#Stellata comes from a species which is no longer in Cattleya
species_filtered <- subset(species_filtered, LowestRank != "Stellata")
valid_tree_tips <- tree$tip.label
species_filtered <- subset(species_filtered, LowestRank %in% valid_tree_tips)

##############################################################

#Extraction of ranks from variable data
# Step 1: Extract the lowest subgeneric rank (series or section)
species_filtered$LowestRank <- sapply(
  strsplit(species_filtered$Subgeneric.Ranks, "\n"),
  function(x) tail(x[x != ""], 1)  # Last non-empty entry
)

# Step 2: Clean up weird or missing entries
species_filtered$LowestRank[species_filtered$LowestRank == ""] <- NA
species_filtered$LowestRank <- trimws(species_filtered$LowestRank)

# View lowest rank assignment
table(species_filtered$LowestRank, useNA = "ifany")


# Create a lookup table
classification_lookup <- data.frame(
  LowestRank = c("Crispae", "Lawrenceanae", "Sophronitis", "Parviflorae", 
                 "Hadrolaelia", "Microlaelia", "Cattleyodes", 
                 "Cattleyella", "Intermediae", "Maximae", "Cattleya"),
  Subgenus = c(rep("Cattleya", 7), "Cattleyella", "Intermediae", "Maximae", "Cattleya")
)

species_filtered <- merge(
  species_filtered,
  classification_lookup,
  by = "LowestRank",
  all.x = TRUE
)

# Check for unmatched ranks
unmatched <- species_filtered[is.na(species_filtered$Subgenus), ]
nrow(unmatched)

############################################################
# Begin building a tree from polytomies of subgeneric ranks

library(ape)

# Load tree from file
tree_lowestrank <- read.tree("../data/cattleya_subgeneric.nwk")

# Step 1: Clean species names (no spaces, safe as tip labels)
species_filtered$SpeciesName <- gsub(" ", "_", species_filtered$Binomial)

# Step 2: Create mapping: rank → species
rank_to_species <- split(species_filtered$SpeciesName, species_filtered$LowestRank)


expanded_tips <- list()

for (rank in names(rank_to_species)) {
  sp <- rank_to_species[[rank]]
  
  if (length(sp) == 1) {
    # Single species: treat as tip with placeholder tree
    expanded_tips[[rank]] <- list(edge = matrix(ncol = 2, nrow = 0),
                                   tip.label = sp,
                                   Nnode = 0)
    class(expanded_tips[[rank]]) <- "phylo"
  } else {
    # Multiple species: make a polytomy (star tree)
    expanded_tips[[rank]] <- stree(length(sp), tip.label = sp, type = "star")
  }
}


# Copy your lowest-rank tree to expand
tree_species <- tree_lowestrank

# Replace each tip with the corresponding mini tree
for (i in seq_along(tree_lowestrank$tip.label)) {
  rank <- tree_lowestrank$tip.label[i]
  
  if (rank %in% names(expanded_tips)) {
    # Expand this tip into species-level subtree
    tree_species <- bind.tree(tree_species, expanded_tips[[rank]], where = which(tree_species$tip.label == rank))
  }
}

# Remove original LowestRank tips from the tree
tree_species <- drop.tip(tree_species, tree_lowestrank$tip.label)

# Plot species level tree
plot(tree_species, cex = 0.5)

# Add hybrid counts
hybrid_counts <- setNames(species_filtered$`X..Descendants`, species_filtered$SpeciesName)

# Match tree to trait data
matched_species <- intersect(tree_species$tip.label, names(hybrid_counts))
tree_species <- drop.tip(tree_species, setdiff(tree_species$tip.label, matched_species))
hybrid_counts <- hybrid_counts[matched_species]

# Visualize
plot(tree_species, cex = 0.5)
tiplabels(pch = 21, bg = heat.colors(length(hybrid_counts))[rank(hybrid_counts)], cex = 1.2)
# Note: red is low value

# Create trait vector: hybrid counts per species
hybrid_counts <- setNames(
  species_filtered$`X..Descendants`,
  species_filtered$SpeciesName
)

# Match species to tree
common_species <- intersect(tree_species$tip.label, names(hybrid_counts))

# Prune tree and trait to match
tree_phylo <- drop.tip(tree_species, setdiff(tree_species$tip.label, common_species))
hybrid_counts <- hybrid_counts[common_species]

library(phytools)

# Phylogenetic tests need branch lengths
#all branches = 1
tree_phylo$edge.length <- rep(1, nrow(tree_phylo$edge))

#grafens method (scaling)
tree_grafen <- compute.brlen(tree_phylo, method = "Grafen")

# Blomberg’s K on species tree (grafen scaling)
K_result <- phylosig(tree_grafen, hybrid_counts, method = "K", test = TRUE)
print(K_result)

# Pagel’s Lambda on species tree (grafen scaling)
lambda_result <- phylosig(tree_grafen, hybrid_counts, method = "lambda", test = TRUE)
print(lambda_result)

#Highly significant pagel
#Low signal for K (not significant)

# Plot tree with grafen method of scaling
plot(tree_grafen, cex = 0.6)



####################################################

# Visualize trait distribution across species tree

tree_clean <- tree_species  
tree_clean <- compute.brlen(tree_clean, method = "Grafen")

# Rebuild abbreviated names
species_filtered$SpeciesAbbrev <- gsub(" ", "_", gsub("^([A-Za-z])[a-z]+\\s+", "\\1. ", species_filtered$Binomial))

# Trait vector using abbreviated names
hybrid_counts <- setNames(species_filtered$`X..Descendants`, species_filtered$SpeciesAbbrev)

# Rename tips in the species-level tree
name_map <- setNames(species_filtered$SpeciesAbbrev, species_filtered$SpeciesName)
tree_clean$tip.label <- name_map[tree_clean$tip.label]

valid_tips <- intersect(tree_clean$tip.label, names(hybrid_counts))
tree_clean <- drop.tip(tree_clean, setdiff(tree_clean$tip.label, valid_tips))
hybrid_counts_clean <- hybrid_counts[tree_clean$tip.label]

# Drop species with NA hybrid counts
keep <- !is.na(hybrid_counts_clean)
hybrid_counts_clean <- hybrid_counts_clean[keep]
tree_clean <- drop.tip(tree_clean, tree_clean$tip.label[!keep])


all(tree_clean$tip.label == names(hybrid_counts_clean))  # should be TRUE

library(phytools)
library(viridis)


contmap_obj <- contMap(tree_clean, hybrid_counts_clean, plot = FALSE)
contmap_obj <- setMap(contmap_obj, viridis(100, option = "D"))
# Format: par(mar = c(bottom, left, top, right))
par(mar = c(5, 4, 4, 2))  # Increase top margin from default 4 to 6
plot(contmap_obj, fsize = 0.6, legend = 0.7, outline = FALSE)


#title 
text(
  x = mean(get("last_plot.phylo", envir = .PlotPhyloEnv)$xx),  # center of x-axis
  y = max(get("last_plot.phylo", envir = .PlotPhyloEnv)$yy) + 2,  # just above highest tip
  labels = "Hybridization Counts Across Cattleya Species Tree",
  font = 2,
  cex = 1.2
)

# Extract coordinates of tips from the last plotted tree
coords <- get("last_plot.phylo", envir = .PlotPhyloEnv)

# Extract x and y coordinates for each tip
x_tip <- coords$xx[1:length(tree_clean$tip.label)]
y_tip <- coords$yy[1:length(tree_clean$tip.label)]

# Add the hybrid count numbers slightly to the right of the tips
text(
  x = x_tip +0.15,  # adjust spacing as needed
  y = y_tip,
  labels = hybrid_counts_clean[tree_clean$tip.label],
  cex = 0.6,
  font = 2,
  adj = 0  # left aligned
)




#Use log count for better visualization

log_counts <- log1p(hybrid_counts_clean)
contmap_obj <- contMap(tree_clean, log_counts, plot = FALSE)
log_counts <- log1p(hybrid_counts_clean)
names(log_counts) <- names(hybrid_counts_clean)  # ensure names match
contmap_obj <- contMap(tree_clean, log_counts, plot = FALSE)
contmap_obj <- setMap(contmap_obj, viridis(100, option = "D"))
par(mar = c(5, 4, 4, 2))  # bottom, left, top, right
plot(contmap_obj, fsize = 0.6, legend = 0.7, outline = FALSE)

library(phylosignal)
library(phylobase)

# Trait vector: must be a named numeric vector
trait_vec <- hybrid_counts_clean[tree_clean$tip.label]

# Build a data.frame (traits must be in data.frame form)
trait_df <- data.frame(hybrids = trait_vec)
rownames(trait_df) <- names(trait_vec)

# Convert to phylo4d object
tree4d <- phylo4d(tree_clean, trait_df)

# test phylogenetic signal on species tree
# blombergs K, Morans I, Abouheifs Cmean
ps <- phyloSignal(tree4d, methods = c("K", "I", "Cmean"), reps = 999)

print(ps)

# K nonsignificant, I and cmean significant (0.001)




